Projection variability: Is the family of sentences really a family?

### 2024-02-14 writing/SuB-paper/projection-operators-SuB.tex
tracking overleaf changes

@@

@@

\usepackage{natbib} 
%\bibpunct{(}{)}{;}{a}{,}{,}
\bibpunct{(}{)}{,}{a}{,}{,}

@@

% drawing
	\usepackage{graphicx}
	\usepackage{xcolor}

@@

%{\large \textbf{Projection variability of clausal complements across different operators}}
{\large \textbf{Projection variability: Is the family of sentences really a family?}}\footnote{We thank Taylor Mahler for assistance in collecting the data presented here as well as valuable comments. We gratefully acknowledge the National Science foundation grant \#1452674 (to Marie-Catherine de Marneffe, Craige Roberts, and Judith Tonhauser), which provided financial support for this project.}\\
Lisa HOFMANN --- \textit{University of Stuttgart}\\
Marie-Catherine DE MARNEFFE --- \textit{UCLouvain}\\
Judith TONHAUSER --- \textit{University of Stuttgart}\\

\textbf{Abstract.} We present experimental evidence that the projection of the content of clausal complements \textbf{(i)} varies across entailment-canceling operators, \textbf{(ii)} that the effect of operator varies between clause-embedding predicates, and \textbf{(iii)} we extend a result of \citealt{degen_are_2022}, that projection ratings in polar questions do not categorically distinguish factive from nonfactive predicates, to cases with negation, the epistemic possibility modal \textit{perhaps}, and conditional antecedents. The observed variability is not captured by existing theoretical accounts of projection (e.g., \citealt{heim_projection_1983,van_der_sandt_presupposition_1992,abrusan_predicting_2011,schlenker_triggering_2021}). Our results suggest that an analysis must consider interactions between predicates and operators and raise important questions for future research on projection.

@@

\section{Introduction} \label{sec:intro}

    Textbooks typically present the family-of-sentences diagnostic for projection without addressing by-operator variation (e.g., \citealt{chierchia_meaning_1990}). For instance, the clausal complement of \emph{discover} is diagnosed as as projective content, if its content (here: that Julian dances salsa) follows not just from \ref{ex:simple}, but also from its variants in \ref{ex:family}, where \ref{ex:simple} has been embedded under an entailment-canceling operator, such as negation \ref{ex:neg}, polar questions \ref{ex:q}, epistemic possibility modals \ref{ex:mod}, or conditional antecedents \ref{ex:cond}.

@@

    % (e.g., \citealt{frege_uber_1892,strawson_referring_1950,kiparsky_fact_1970,karttunen_conventional_1979}).

@@

	\ex. \label{ex:family}
		\a. \label{ex:neg}
			{\bf Negation:} \hfill
			\emph{Cole didn't discover that Julian dances salsa.}
		\b. \label{ex:q}
			{\bf Polar Question:} \hfill
			\emph{Did Cole discover that Julian dances salsa?}
		\c. \label{ex:mod}
			{\bf Modal:} \hfill
			\emph{Perhaps Cole discovered that Julian dances salsa.}
		\d. \label{ex:cond}
			{\bf Conditional:} \hfill
			\emph{If Cole discovered that Julian dances salsa, Logan will be joyful.}
		\z.
	\z.

    However, prior research is suggestive of variation in the effect of different entailment-cancelling operators on projection. For instance, \citet{karttunen_observations_1971} proposed a distinction between English factive predicates (e.g., \textit{regret}) and semi-factives (e.g., \textit{discover}).
    Based on the examples in \Next, he suggested that the CC of true factives projects across all four operators, but that of semi-factives does not always project from under polar questions, modals, or conditionals.\footnote{See \citealt{beaver_have_2010} for naturally occurring examples in which the CCs of factive predicates do not project from under the four operators, thereby calling into question the distinction between factive and semi-factive predicates. For experimental research on the distinction see \citealt{djaerv-etal2016} and \citealt{djarv_cognitive_2018}. We return to this literature in section \ref{s:analysis}.}

@@

    There have been two experimental investigations of by-operator projection variation. First, the experiment reported in \citet{smith_relationship_2014} investigated projection from under negation and antecents of conditionals for various types of English projective contents. They found that the expressive content of epithets (e.g., \textit{idiot}) and the CC of \textit{know} was more projective under negation than conditionals. In contrast, the content of appositive relative clauses and the preparatory content of \textit{win} showed the opposite pattern, and the existential presupposition of clefts showed no difference. %However, their task involved asking participants how surprised they would be to learn the content under investigation after observing the utterance. Since other factors besides speaker commitment modulate discourse (un)expectedness (see e.g., \citealt{zimmermann_grammatical_2011,tonnis_german_2021}), it is unclear that they (only) measured projection.

    Second, \citet{sieker_projective_2022}  compared the projection of German factives (\textit{wissen} `know', \textit{bereuen} `regret', \textit{ent\-hüllen} `reveal') and semi-factives (\textit{bemerken} `notice', \textit{entdecken} `discover', \textit{herausfinden} `find out') between the four operators in \ref{ex:family}. The results showed that the CCs of the predicates investigated projected more from under negation than the other three operators. contrary to what \citealt{karttunen_observations_1971} suggested, a comparison of factive and semifactive predicates did not reveal that the CCs of factive predicates project more from under polar questions, modals, or antecedents of conditionals than the CCs of nonfactive predicates.

@@

    This paper reports on the results of a set of experiments that were designed to compare projection from under the four entailment-canceling operators in \ref{ex:family}. Our experiments extend the empirical scope of prior research by investigating projection for a broader set of contents, namely the contents of the complements of the 20 English clause-embedding predicates in \ref{ex:predicates}:

    \ex. \label{ex:predicates}
    		\a. (Semi)factive predicates: {\em be annoyed, know, reveal, discover, see}
      \b. Nonfactive predicates: {\em acknowledge, admit, announce, confess, confirm, establish, hear, inform, prove, be right, demonstrate, pretend, say, suggest, think}

    As shown, the  20 predicates include both (semi)factive and nonfactive predicates. The inclusion of the latter set of predicates is motivated by the results of the empirical investigations in \citealt{degen_are_2022}, which suggested that the contents of complements of nonfactive predicates is also projective and that projection does not categorically distinguish factive and nonfactive predicates. As \poscite{degen_are_2022} result was established based on sentences where the predicates were embedded under polar questions, our experiments, which feature all four entailment-canceling operators, serve to replicate and extend this result of \citealt{degen_are_2022}.

@@

\section{Experiments}

	To assess the effect of entailment-cancelling operator and clause-embedding predicate on projection, we collected projection judgments for the CCs of the 20 clause-embedding predicates in four sets of experiments.\footnote{The data and analysis scripts can be found in the following GitHub repository: \url{https://github.com/judith-tonhauser/CommitmentBankPlus}. {\bf LETS BE MORE SPECIFIC HERE ABOUT WHERE THINGS RELATED TO THIS PAPER ARE}} The predicates were embedded under negation in Exps.~1, under polar questions in Exps.~2, under the epistemic possibility modal {\em perhaps} in Exps.~3, and in conditional antecedents in Exps.~4. Each set of experiments consisted of three experiments that differed in the at-issueness measure used in a separate block. In this paper, we limit our attention to the projection ratings collected in the 12 experiments.

	Projection was measured in each experiment with the `certain that' diagnostic, which has been used to measure projection with both polar interrogative and declarative sentences (see, e.g., \citealt{tonhauser_prosodic_2016,djarv_prosodic_2017,stevens_rational_2017,lorson_influence_2018,tonhauser_how_2018,mahler_does_2019,mahler_social_2020,de_marneffe_commitmentbank_2019}).\footnote{For other diagnostics of projection see, e.g.\ \citealt{smith_projection_2011,xue_correlation_2011}, and \citealt{tonhauser_toward_2013}; and discussion in \citealt{tonhauser_how_2018}.} This diagnostic was also used in \citealt{sieker_projective_2022}. Under this diagnostic, participants are presented with utterances like those in \Next. They are asked to rate whether the (named) speaker is certain of the CC.

	\ex. \label{ex:certain-that}
		\a. Christopher: \emph{\lq Cole didn't discover that Julian dances salsa.\rq}
		\b. Christopher: \emph{\lq Did Cole discover that Julian dances salsa?\rq}
		\c. Christopher: \emph{\lq Perhaps Cole discovered that Julian dances salsa.\rq}
		\d. Christopher: \emph{\lq If Cole discovered that Julian dances salsa, Logan will be joyful.\rq}
		\z.
		Projection question: Is Christopher certain that Julian dances salsa?
	\z.

    We assume, following \citealt{tonhauser_how_2018} and \citealt{degen_are_2022}, that judgments of speaker certainty about some propositional content reflect speaker commitment to that content, that is, projection. If a participant interprets an utterance like \Last[a--d] in a way that the speaker (e.g., Christopher) is certain of the CC, it is assumed to project out of the entailment-canceling environment. If a participant does not take the speaker to be certain of the CC, this is taken to indicate that it does not project.

@@

		\subsubsection{Participants}
			We recruited 250-300 participants for each of the 12 experiments. Participants for Exp.~1q were recruited on Amazon's Mechanical Turk platform; the participants were required to have U.S.\ IP addresses and at least 99\% of previously approved HITs. Participants for the remaining experiments were recruited on Prolific; these participants were required to reside in the US, to be born in the US, to have English as their first language, and to have an approval rating of at least 99\%. See Supplement \ref{app:d-participants} for further information about the participants.

@@

            The target sentences consisted of 400 combinations of the 20 clause-embedding predicates in \ref{ex:predicates} with 20 embedded clauses (provided in Supplement \ref{app:a-clauses}). As mentioned above, the predicates were embedded under negation in Exps.~1, under polar questions in Exps.~2, under the epistemic possibility modal \emph{perhaps} in Exps.~3, and in conditional antecedents in Exps.~4, for a total of 400 target stimuli in each of the four sets of experiments.

			%We tested the 20 clause-embedding predicates in \Next, which have shown projection variability in polar questions (\citealt{degen_are_2022}). They comprise three factive predicates \Next[a], two semi-factives \Next[b], nine that have been characterized as optionally factive in \citealt{kiparsky_fact_1970} \Next[c], and six non-factives which include two veridical predicates \Next[d], and four non-veridical ones \Next[e].

	%		\ex. 20 clause-embedding predicates \hfill \citealt{degen_are_2022}
			%	\a. canonically factive:					\emph{be annoyed, know, reveal}
				%\b. semi-factive: 		\emph{discover, see}
%				\b. optionally factive: \newline \emph{acknowledge admit, announce, confess, confirm, establish, hear, inform, prove}
%				\b. Non-factive veridical:
%					\emph{be right, demonstrate}
%				\b. Non-factive non-veridical:
%				\emph{pretend, say, suggest, think}
%				\z.
%			\z.

To assess whether participants were attending to the task, each experiment included six control stimuli. For details on the six control stimuli see Supplement \ref{app:c-control}.

@@

			Participants were told to imagine that they are at a party and that, on walking into the kitchen, they overhear somebody say something to somebody else.
            %
            On each trial, participants read an utterance and responded to the `certain that'   question. They gave their response on a slider marked `no' (coded as 0) at one end and and `yes' (coded as 1) on the other. A sample trial from Exps.~1 is illustrated in Figure~\ref{fig:trial}. Following, e.g., \citet{tonhauser_how_2018}, higher ratings of speaker certainty could reflect one of two things:
            %
            On the one hand, higher ratings could reflect greater speaker commitment towards the CC, and therefore greater projection. This assumes that projection (and the relevant notion of speaker commitment) can itself be gradient.
            %
            Conversely, higher ratings could also reflect the probability that a participant associates with a projective interpretation for the utterance. Under this interpretation, commitment may be interpreted categorically, and the gradience is a result of the uncertainty about possible interpretations (see also discussion in \citealt{grove_factivity_2023}, which suggests the latter kind of interpretation).

@@

			\begin{figure}[ht]
				\centering
				\fbox{\includegraphics[width = .7\linewidth]{task-1n-proj.eps}}
				\caption{A sample trial in Exps.~1. In the corresponding trials in Exps.~2-4, participants were presented with an utterance with a different entailment-cancelling operator.}
				\label{fig:trial}
			\end{figure}

@@

    We first address the question of whether there is by-operator variation (section \ref{s:by-op}) and then the question of whether there is by-predicate variation in the observed by-operator variation (section \ref{s:by-pred}).

@@

        Fig.~\ref{fig:op-ratings} shows the mean certainty ratings by entailment-canceling operator, aggregating over the clause-embedding predicates. As shown, there is projection variability by operator: The CCs of the clause-embedding predicates were relatively more projective when embedded in the antecedent of a conditional than in a polar interrogative, where they were relative more projective than when they were embedded under negation or the epistemic modal {\em perhaps}.

@@

   These generalizations are supported by Model \#1 reported in Table~\ref{t:op-model}.

@@

			Like \citet{sieker_projective_2022}, we observe a small main effect of entailment-cancelling operator on projection. However, in contrast to their result that projection is strongest from under negation, our results suggest that projection is strongest from antecedents of conditionals.
            We note, however, that the differences in mean certainty ratings between the four entailment-canceling operators are very small. This suggests that, when abstracting away from individual projective contents, projection from under the four entailment-canceling operators is very similar. In other words, when abstracting away from individual projective contents, the family of sentences really are a family.

@@

  Fig.~\ref{fig:op-pred-ratings} shows mean certainty ratings by predicate and entailment-canceling operator (aggregating over the 20 CCs that each predicate was paired with). As shown, there is by-operator variation for the CCs of all 20 clause-embedding predicate, that is, there is not a single predicate for which there is no by-operator variation. The observed variation does not align with widely assumed predicate classes. For instance, the five (semi)factive predicates exhibit five different patterns: 
i) the CC of \emph{be annoyed} projects more from under negation and questions than conditionals and modal {\em perhaps}; ii) the CC of \emph{know} projects most from under questions, less from questions and negation, and least from under the modal {\em perhaps}; iii) the CC of \emph{see} projects most from under questions and conditionals, less from under negation, and least from under the modal \emph{perhaps}; iv) the CC of  \emph{discover} projects more from under conditionals and questions than negation and modal {\em perhaps}; and v) the CC of \emph{reveal} projects most from conditionals, less from questions, and least from negation and modal \emph{perhaps}. Thus, contrary to what \citealt{karttunen_observations_1971} proposed, there is no predicate whose CC projects uniformly from under all four entailment-canceling operators (as he suggested for factive predicates) and the purported semi-factive predicates  \emph{discover} and \emph{reveal} do not project more from under negation than the other three entailment-canceling operators. We also observe by-operator projection variation for nonfactive predicates. Some of this variation aligns with that observed for factive predicates: For instance, the CC of \emph{inform} exhibits the same pattern as the CC of \emph{know}, and the CC of \emph{acknowledge} exhibit the same pattern as that of the CC of \emph{discover}. Other nonfactive predicates exhibit other patterns: For instance, the CCs of \emph{admit}, \emph{confess} and \emph{announce} project most from the antecedents of conditionals than the other three operators.

\begin{figure}[ht]
				\centering
				\includegraphics[width = \linewidth]{predicate-operator-graph-1}
				\caption{Mean certainty ratings by predicate and operator with 95\% bootstrapped confidence intervals. Predicates are ordered by mean certainty rating pooled over the four operators.
Embedding operator coded by letter and color:  \texttt{N} (light blue): negation, \texttt{M} (orange): modals, \texttt{C} (green): conditional antecedents, \texttt{Q} (dark blue): polar questions.}
				\label{fig:op-pred-ratings}
			\end{figure}

The observation that there is by-operator variation for the CC of each of the 20 clause-embedding predicates is confirmed by the statistical analysis. We fitted...

  The CC of \emph{know} projects less from under negation than questions, but more from under negation than modals, while the difference between negation and conditionals is not significant.
			%
			These generalizations are supported by models\ \# 2--4 in Table\ \ref{t:models}, which also each have at least $34$ highly significant interaction terms (out of $57$ possible interactions of operator and predicate).

			\begin{table}[ht]
				\centering
				\begin{tabular}{llrrrr}
					Model & & Estimate & Std. Error & t-value\\
					\midrule
					\#2 & Intercept: \emph{\bf be annoyed}/negation & 0.87 & 0.01 & 79.86 & ***\\
					& operator: conditional & -0.12 & 0.02  & -7.36 & ***\\
					& operator: modal & -0.16 & 0.02  & -10.01 & ***\\
					& operator: question & 0.02 & 0.01 & 1.72 & n.s.\\
					\midrule
					\#3 & Intercept: \emph{\bf discover}/negation & 0.68 & 0.01 & 62.70 & ***\\
					& operator: conditional & 0.11 & 0.02 & 7.11 & ***\\
					& operator: modal & -0.06 & 0.02 & -3.63 & ***\\
					& operator: question & 0.10 & 0.01 & 7.08 & ***\\
					\midrule
					\#4 & Intercept: \emph{\bf know}/negation & 0.79 & 0.01 & 72.97 & ***\\
					& operator: conditional & 0.00 & 0.02 & -0.06 & n.s.\\
					& operator: modal & -0.14 & 0.02 & -9.18 & ***\\
					& operator: question & 0.08 & 0.01 & 5.67 & ***\\
					\bottomrule
				\end{tabular}
				\caption{\small Excerpt of the output from three linear mixed effects models, with fixed effects: operator, predicate, and their interaction; random effect: participant intercepts.
				Models were fit with \texttt{lme4, lmertest} in \texttt{R}. Models \textbf{\#2--4} also had at least $34$ highly significant interaction terms of \texttt{operator} and \texttt{predicate} with $p < 0.001$ (not shown here).\label{t:models}}
			\end{table}

Our results align with those of \citealt{smith_relationship_2014}, who also observed by-expression by-operator variation. However, while they found that the CC of \emph{know} projects more from under negation than the antecedent of a conditional, we did not find an effect here. We hypothesize that this difference is due to the difference in projection diagnostic used.

>>>>

\usepackage{subcaption}

@@

\newcommand{\mcdm}[1]{\textbf{\color{blue}[mcdm: #1]}}

@@

\usepackage[longnamesfirst]{natbib} 
%\bibpunct{(}{)}{;}{a}{,}{,}
\bibpunct{(}{)}{,}{a}{,}{,}

@@

% drawing
	\usepackage{graphicx}
	\usepackage{xcolor}
    \definecolor{op-n-color}{HTML}{56B4E9}
    \definecolor{op-m-color}{HTML}{E69F00}
    \definecolor{op-c-color}{HTML}{009E73}
    \definecolor{op-q-color}{HTML}{0072B2}
    \definecolor{pred-fcv-color}{HTML}{CC79A7}

@@

{\large \textbf{Projection variability: Is the family of sentences really a family?}}\footnote{We thank Taylor Mahler for assistance in collecting the data presented here as well as valuable comments. We gratefully acknowledge the National Science foundation grant \#1452674 (to Marie-Catherine de Marneffe, Craige Roberts, and Judith Tonhauser), which provided financial support for this project.}\\
Lisa HOFMANN --- \textit{University of Stuttgart}\\
Marie-Catherine DE MARNEFFE --- \textit{UCLouvain}\\
Judith TONHAUSER --- \textit{University of Stuttgart}\\

\textbf{Abstract.} We present experimental evidence \textbf{(i)} that the projection of the content of clausal complements  varies across entailment-canceling operators, \textbf{(ii)} that the effect of operator varies between clause-embedding predicates, and \textbf{(iii)} we extend a result of \citealt{degen_are_2022}, that projection ratings in polar questions do not categorically distinguish factive from non-factives predicates, to cases with negation, the epistemic possibility modal \textit{perhaps}, and conditional antecedents. The observed variability is not captured by existing theoretical accounts of projection (e.g., \citealt{heim_projection_1983,van_der_sandt_presupposition_1992,abrusan_predicting_2011,schlenker_triggering_2021}). Our results suggest that an analysis must consider interactions between predicates and operators and raise important questions for future research on projection.

@@

\section{Introduction} \label{sec:intro}
    The so-called `family of sentences' diagnostic is the standard way of diagnosing whether a content is projective in formal semantics/pragmatics (e.g., \citealt{chierchia_meaning_1990}). For instance, the clausal complement of \emph{discover} is diagnosed as as projective content, if its content (here: that Julian dances salsa) follows not just from \ref{ex:simple}, but also from the variants in \ref{ex:family}, where \ref{ex:simple} is embedded under an entailment-canceling operator, such as negation \ref{ex:neg}, a polar question \ref{ex:q}, an epistemic possibility modal \ref{ex:mod}, or in a conditional antecedent \ref{ex:cond}.

@@

@@

	\ex. \label{ex:family}
		\a. \label{ex:neg}
			{\bf Negation:} \hfill
			Cole didn't discover that Julian dances salsa.
		\b. \label{ex:q}
			{\bf Polar Question:} \hfill
			Did Cole discover that Julian dances salsa?
		\c. \label{ex:mod}
			{\bf Modal:} \hfill
			Perhaps Cole discovered that Julian dances salsa.
		\d. \label{ex:cond}
			{\bf Conditional:} \hfill
			If Cole discovered that Julian dances salsa, Logan will be joyful.
		\z.
	\z.

    Some research, however, suggests that entailment-cancelling operators may affect projection differentially. For instance, \citet{karttunen_observations_1971} proposed distinguishing English factive predicates (e.g., \textit{regret}) from semi-factives (e.g., \textit{discover}). Based on \Next, he suggested that the content of the complement (CC) of true factives consistently projects across the operators in \Last, but that of semi-factives does not always project from under polar questions, modals, or conditionals.%

@@

    There have been two experimental investigations of by-operator projection variation. First, \citet{smith_relationship_2014} investigated projection from under negation and conditional antecedents for various types of English projective contents. They found that the expressive content of epithets (e.g., \textit{idiot}) and the CC of \textit{know} was more projective under negation than conditionals. In contrast, the content of appositive relative clauses and the preparatory content of \textit{win} showed the opposite pattern, and the existential presupposition of clefts showed no difference.
    %However, their task involved asking participants how surprised they would be to learn the content under investigation after observing the utterance. Since other factors besides speaker commitment modulate discourse (un)expectedness (see e.g., \citealt{zimmermann_grammatical_2011,tonnis_german_2021}), it is unclear that they (only) measured projection.

    Second, \citet{sieker_projective_2022}  compared the projection of the CCs of German factives (\textit{wissen} `know', \textit{bereuen} `regret', \textit{ent\-hüllen} `reveal') and semi-factives (\textit{bemerken} `notice', \textit{entdecken} `discover', \textit{herausfinden} `find out') from under the four operators in \ref{ex:family}. The results suggest that the CCs project more from under negation than from under the other three operators. Contrary to what \citet{karttunen_observations_1971} suggested, a comparison of the factive and semi-factive predicates did not reveal that the CCs of factive predicates project more from under polar questions, modals, or antecedents of conditionals than the CCs of semi-factive predicates.\footnote{The factive/semi-factive distinction is also called into question by naturally occurring examples where the CCs of factive predicates do not project from under the four operators (see \citealt{beaver_have_2010,de_marneffe_did_2012,de_marneffe_commitmentbank_2019}). For experimental research on the distinction, see \citealt{djaerv-etal2016} and \citealt{djarv_cognitive_2018}.
    }

@@

    This paper reports on the results of a set of experiments that were designed to compare projection from under the four entailment-canceling operators in \ref{ex:family} in English. Our experiments extend the empirical scope of prior research on by-operator projection variation by investigating projection for a larger set of contents, namely the contents of the complements of the 20 English clause-embedding predicates in \ref{ex:predicates}, from \citet{degen_are_2022}.

    \ex. \label{ex:predicates}
    		\a. (Semi-)factive predicates: {\em be annoyed, know, reveal, discover, see}
      \b. Non-factive predicates: {\em acknowledge, admit, announce, confess, confirm, establish, hear, inform, prove, be right, demonstrate, pretend, say, suggest, think}

    The five predicates in \Last[a] have been characterized as factive or semi-factive. Our set of predicates also includes the 15 non-factives in \Last[b]. Including non-factive predicates in investigations of projection is motivated by the empirical investigations in \citealt{de_marneffe_commitmentbank_2019} and \citealt{degen_are_2022}, which suggest that the CCs of non-factives may also project and that projection ratings do not categorically distinguish factive and non-factive predicates.

    The results of our study suggest that the projection of the CCs of these 20 predicates is affected differently by the four entailment-cancelling operators in \ref{ex:family},  but not in a way that is consistent with \poscite{karttunen_observations_1971} factive/semi-factive distinction. The results also replicate a result of \citealt{degen_are_2022}, namely that projection from under polar questions does not categorically distinguish factive and non-factive predicates. Finally, the results of our experiment extend this result of \citealt{degen_are_2022} to projection from under the other three entailment-canceling operators, thereby solidifying their claim that ``research on projective content has a much broader empirical scope than previously assumed'' (p.585), as this scope includes the CCs of both factive and non-factive predicates.

@@

\section{Experiments}\label{s2}

	To assess the effect of entailment-cancelling operator and clause-embedding predicate on projection, we collected projection judgments for the CCs of the 20 clause-embedding predicates in four sets of experiments.
    %
    The predicates were embedded under negation in Exps.~1, under polar questions in Exps.~2, under the epistemic possibility modal {\em perhaps} in Exps.~3, and in conditional antecedents in Exps.~4. Each set of experiments contained three experiments, that differed in an at-issueness measure used in a separate block. In this paper, we limit our attention to the projection ratings collected in these twelve experiments.\footnote{\label{footnote-supplement}The experiments, data and analysis scripts, as well as supplements referred to in this paper can be found in the following GitHub repository: \url{https://github.com/judith-tonhauser/CommitmentBankPlus}.}

	In all twelve experiments, projection was measured with the `certain that' diagnostic, which has been used to measure projection with both polar interrogative and declarative sentences (see, e.g., \citealt{tonhauser_prosodic_2016,djarv_prosodic_2017,stevens_rational_2017,lorson_influence_2018,tonhauser_how_2018,mahler_does_2019,mahler_social_2020,de_marneffe_commitmentbank_2019}).\footnote{For other diagnostics of projection see, e.g.\ \citealt{smith_projection_2011,xue_correlation_2011}, and \citealt{tonhauser_toward_2013}; and discussion in \citealt{tonhauser_how_2018}.} This diagnostic was also used in \citealt{sieker_projective_2022}. Under this diagnostic, participants are presented with utterances like those in \Next, and asked to rate whether the (named) speaker is certain of the CC.

	\ex. \label{ex:certain-that}
		\a. Christopher: ``Cole didn't discover that Julian dances salsa.''
		\b. Christopher: ``Did Cole discover that Julian dances salsa?''
		\c. Christopher: ``Perhaps Cole discovered that Julian dances salsa.''
		\d. Christopher: ``If Cole discovered that Julian dances salsa, Logan will be joyful.''
		\z.
		Projection question: Is Christopher certain that Julian dances salsa?
	\z.

    We assume, following \citet{tonhauser_how_2018} and \citet{degen_are_2022}, that judgments of speaker certainty about the embedded content reflect speaker commitment to that content, that is, projection. If a participant interprets utterances like \Last[a--d] in a way that the speaker (here, Christopher) is certain of the CC, the CC is assumed to project. If a participant does not take the speaker to be certain of the CC, the CC is taken to not project.

@@

		\subsubsection{Participants}
			We recruited 250-300 participants for each of the 12 experiments. Participants for one experiment were recruited on Amazon's Mechanical Turk platform. These participants were required to have U.S.\ IP addresses and at least 99\% of previously approved HITs. Participants for the remaining experiments were recruited on Prolific. These participants were required to reside in the US, to be born in the US, to have English as their first language, and to have an approval rating of at least 99\%. See Supplement \ref{app:d-participants} (in the repository linked to in footnote \ref{footnote-supplement}) for further information about the participants.

            \begin{figure}[h!]
				\centering
				\fbox{\includegraphics[width = .6\linewidth]{task-1n-proj.eps}}
				\caption{A sample trial in Exps.~2. In the corresponding trials in the other experiments, participants were presented with an utterance with a different entailment-cancelling operator.}
				\label{fig:trial}
			\end{figure}

@@

            The target sentences consisted of the 400 combinations of the 20 clause-embedding predicates in \ref{ex:predicates} with 20 embedded clauses (provided in Supplement \ref{app:a-clauses}). As mentioned above, the predicates were embedded under negation in Exps.~1, under polar questions in Exps.~2, under the epistemic possibility modal \emph{perhaps} in Exps.~3, and in conditional antecedents in Exps.~4, for a total of 400 target stimuli in each of the four sets of experiments. To assess whether participants were attending to the task, each experiment included six control stimuli. For details on the six control stimuli, see Supplement \ref{app:c-control}.

@@

			Participants were told to imagine that they are at a party and that, when walking into the kitchen, they overhear somebody say something to somebody else.
            %
            On each trial, participants read an utterance and responded to the `certain that'   question. They gave their response on a slider marked `no' (coded as 0) at one end and `yes' (coded as 1) on the other. A sample trial from Exps.~1 is given in Figure~\ref{fig:trial}.

            Following \citet{tonhauser_how_2018}, higher ratings of speaker certainty could reflect one of two things:
            %
            On the one hand, higher ratings could reflect greater speaker commitment towards the CC, and therefore greater projection. This assumes that speaker commitment is gradient.
            %
            Conversely, higher certainty ratings could reflect the probability that an interpreter takes the speaker to be committed to the CC.  On this interpretation, speaker commitment may be a binary, categorical property and projection variation is a result of the uncertainty about possible interpretations. %, which suggests the latter kind of interpretation). 
            In this paper, we remain agnostic about the underlying interpretation of projectivity as a gradient property, though our discussion of projection variability will be in line with the first interpretation (for discussion, see also \citealt{grove_factivity_2023}).

@@

@@

    We first address the question of whether there is by-operator variation (section \ref{s:by-op}) and then the question of whether there is by-predicate variation in the observed by-operator variation (section \ref{s:by-pred}). Finally, in section \ref{s:factivity}, we relate the results to the result of \citealt{degen_are_2022}, that projection from under polar questions does not categorically distinguish factive and non-factive predicates.

@@

        \mcdm{again needs to be consistent, sometimes Figure} Figure~\ref{fig:op-ratings} shows the mean certainty ratings by entailment-canceling operator, aggregating over the clause-embedding predicates. As shown, there is projection variability by operator: The CCs of the clause-embedding predicates were relatively more projective when embedded in the antecedent of a conditional than in a polar interrogative, where they were relatively more projective than when they were embedded under negation or the epistemic modal {\em perhaps}.

@@

   These generalizations are supported by Model \#1 reported in Table~\ref{t:op-model}.\mcdm{we don't know what Model \#1 is}

@@

			Like \citet{sieker_projective_2022} for German, we observe a small main effect of entailment-cancelling operator on projection. However, in contrast to their result that projection is strongest from under negation, our results suggest that projection is strongest from antecedents of conditionals.\mcdm{does it make sense to compare different predicates?}
            We note, however, that the differences in mean certainty ratings between the four entailment-canceling operators are very small. This suggests that, when abstracting away from individual projective contents, projection from under the four entailment-canceling operators is very similar. In other words, when abstracting away from individual projective contents, the family of sentences really are a family.

@@

    Figure~\ref{fig:op-pred-ratings} shows mean certainty ratings by entailment-canceling operator for the five (semi-)factive predicates (indicated in pink), and the 15 non-factive predicates, where predicates are ordered by their overall mean certainty ratings.

    \begin{figure}[ht]
        \centering
		\includegraphics[width = \linewidth]{predicate-operator-graph-1}
		\caption{Mean certainty ratings by predicate and operator with 95\% bootstrapped confidence intervals. \textcolor{pred-fcv-color}{(Semi-)factive} and non-factive predicates are ordered by mean certainty rating pooled over the four operators.
        Embedding operator coded by letter and color:  \textcolor{op-n-color}{\texttt{N}: negation}, \textcolor{op-m-color}{\texttt{M}: modal \emph{perhaps}}, \textcolor{op-c-color}{\texttt{C}: conditional antecedents}, \textcolor{op-q-color}{\texttt{Q}: polar questions}.}
		\label{fig:op-pred-ratings}
	\end{figure}

    As shown, there is by-operator variation for the CCs of all 20 clause-embedding predicate, that is, there is not a single predicate for which there is no by-operator variation. Further, the effect of operator differs between predicates. Examining the differences in certainty-ratings between the predicates for the four operators suggests that the observed variation does not align with widely assumed predicate classes. For instance, the five (semi-)factive predicates exhibit five different patterns:

    i) the CC of \emph{be annoyed} projects more from under negation and questions than conditionals and modal {\em perhaps}; ii) the CC of \emph{know} projects most from under questions, less from conditionals and negation, and least from under the modal {\em perhaps}; iii) the CC of \emph{see} projects most from under questions and conditionals, less from under negation, and least from under the modal \emph{perhaps}; iv) the CC of  \emph{discover} projects more from under conditionals and questions than negation and modal {\em perhaps}; and v) the CC of \emph{reveal} projects most from conditionals, less from questions, and least from negation and modal \emph{perhaps}. Thus, contrary to what \citealt{karttunen_observations_1971} proposed, there is no predicate whose CC projects uniformly from under all four entailment-canceling operators (as he suggested for factive predicates) and the purported semi-factive predicates  \emph{discover} and \emph{reveal} do not project more from under negation than the other three entailment-canceling operators. We also observe by-operator projection variation for non-factives predicates. Some of this variation aligns with that observed for factive predicates: For instance, the CC of \emph{inform} exhibits the same pattern as the CC of \emph{know}, and the CC of \emph{acknowledge} exhibit the same pattern as that of the CC of \emph{discover}. Other non-factives predicates exhibit other patterns: For instance, the CCs of \emph{admit}, \emph{confess} and \emph{announce} project most from the antecedents of conditionals than the other three operators.

    These generalizations are confirmed by the statistical analysis. We fitted a Bayesian mixed-effects beta regression to the subset of the data containing the observations for each predicate, using the \texttt{brms} package (\citealt{burkner_brms_2017}) in R (\citealt{r_core_team_r_2014}). Each of the 20 models used low-information priors and estimated the effect of the \lq operator\rq\ (fixed effect) on scaled\footnote{The beta regression model assumes a beta-distribution defined over the open unit interval $(0,1)$, excluding boundaries. Following \citet{degen_are_2022}, we scaled the response variable, which is distributed over the closed interval $[0,1]$ (including boundaries), using the method in \citealt{smithson_better_2006}.} projection ratings, with random by-item intercepts (where an item is a complement clause), and by-item slopes for operators.

    For each predicate, the differences between operators were established by a pairwise comparison of posterior samples for the four operators using the \texttt{emmeans} package (\citealt{lenth_emmeans_2024}). This comparison yielded the $95\%$ high density interval (HDI) of estimated marginal mean differences for each pairwise combination of operators. We assume that two conditions differ when their HDI does not include $0$.

    The results are indicated in Figure~\ref{fig:op-pred-analysis}, which shows shows mean certainty ratings by entailment-canceling operator for the 20 predicates, here grouped by predicate. As in Figure~\ref{fig:op-pred-ratings}, (semi-)factive predicates are indicated in pink, and predicates are ordered by their overall mean certainty ratings. Non-zero differences between two operators are indicated by lines spanning the two operators below each predicate facet.

    \begin{figure}[t]
        \centering
		\includegraphics[width = \linewidth]{analysis.pdf}
		\caption{Mean certainty ratings by predicate (\textcolor{pred-fcv-color}{(semi-)factive}, non-factive) and operator (\textcolor{op-m-color}{modal},  \textcolor{op-n-color}{negation}, \textcolor{op-q-color}{polar questions}, \textcolor{op-c-color}{conditional antecedents}) with 95\% bootstrapped confidence intervals. Violin plots indicate kernel probability density of observations. Below each facet, a line spanning two operators indicates a non-zero difference according to the pairwise comparison of operators. The linetype indicates whether the difference is
        at least $1$ (solid line: —),
        below $1$ but at least $0.5$ (dashed line: – –),
        or below $0.5$ but above $0$ (dotted line: . . . )}
		\label{fig:op-pred-analysis}
	\end{figure}

    Because we found statistically robust non-zero differences by-operator differences for each of the predicates, and these differ beween predicates, the analysis also supports the more general claims that there is by-operator variation for the CC of each of the 20 clause-embedding predicates, and that the effect of operator differs between predicates.

    As a result, our findings align with those of \citealt{smith_relationship_2014}, who also observed by-expression variation in the effect of operator. However, while they found that the CC of \emph{know} projects more from under negation than the antecedent of a conditional, we did not find an effect here. We hypothesize that this difference is due to the difference in projection diagnostic used.
